---
title: "Final Assignment"
output: pdf_document
date: "2025-08-11"
editor_options: 
  chunk_output_type: inline
---



Loading libraries

```{r}
library(ggplot2)
```



```{r}
# Loading gene expression data.

gene_expression <- read.csv2(paste0("C:/Users/Guest1/Dropbox/",
                                    "VuiFinal/QBS103_GSE157103_genes.csv"),sep = ",")

# Adding X as row.names of the dataframe.

rownames(gene_expression) <- gene_expression$X

# Removing first column.

gene_expression <- gene_expression[,-c(1)]

# Making gene expression values numeric.

gene_expression_num <- data.frame(apply(gene_expression,2,as.numeric))

rownames(gene_expression_num) <- rownames(gene_expression)

gene_expression <- gene_expression_num

dim(gene_expression)

# Transposing gene expression dataframe.

gene_expression_t <- as.data.frame(t(gene_expression))

# Assigning column and row names.

colnames(gene_expression_t) <- rownames(gene_expression)

rownames(gene_expression_t) <- colnames(gene_expression)

# Reading covariates data.

covariates_data <- read.csv2(paste0("C:/Users/Guest1/Dropbox/",
                                    
          "VuiFinal/QBS103_GSE157103_series_matrix-1.csv"),sep = ",")

# Selecting covariates

# Selecting three continuous and three categorical covariates.

continuous_cov <- c("hospital.free_days_post_45_day_followup","ferritin.ng.ml.",
                    "fibrinogen")

# Selecting three categorical covariates.

categorical_cov <- c("sex","mechanical_ventilation","disease_status")

# Subseting original covariates dataframe.

covariates_data_filt <- covariates_data[,c(continuous_cov,categorical_cov)]
```

# Selecting genes.

To apply a criteria to select genes for this assignment, in this section I looked for the genes that presented the strongest associations with the selected categorical and continuous variables. For continuous variables perason´s correlations were computed and for categorigal t-tests were carried out. 

```{r, warning=FALSE}
# Selecting genes with top accocititon to our covariates.

# Computing correlations with continuos covariates  

# Getting the top associated genes for hospital free days post 45 days follow up

correlations <- c()

for (i  in 1:nrow(gene_expression)){
  
  cor_results <- cor.test(as.numeric(gene_expression[i,]), 
                          as.numeric(covariates_data_filt[,1]), method = "pearson")
  
  correlations <- c(correlations,cor_results$estimate)
}


rownames(gene_expression)[which.max(correlations)]

max(correlations,na.rm = TRUE)

rownames(gene_expression)[which.min(correlations)]

min(correlations,na.rm = TRUE)

# ABCD4 top pos corr
# ABCB6 topneg corr

# Getting the top associated genes for ferritin

correlations <- c()

for (i  in 1:nrow(gene_expression)){
  
  cor_results <- cor.test(as.numeric(gene_expression[i,]), 
                          as.numeric(covariates_data_filt[,2]), method = "pearson")
  
  correlations <- c(correlations,cor_results$estimate)
}


rownames(gene_expression)[which.max(correlations)]

max(correlations,na.rm = TRUE)

rownames(gene_expression)[which.min(correlations)]

min(correlations,na.rm = TRUE)

# A4GALT top pos corr
# AAMP topneg corr

# Getting the top associated genes for fibrinogen

correlations <- c()

for (i  in 1:nrow(gene_expression)){
  
  cor_results <- cor.test(as.numeric(gene_expression[i,]), 
                          as.numeric(covariates_data_filt[,3]), method = "pearson")
  
  correlations <- c(correlations,cor_results$estimate)
}


rownames(gene_expression)[which.max(correlations)]

max(correlations,na.rm = TRUE)

rownames(gene_expression)[which.min(correlations)]

min(correlations,na.rm = TRUE)

# ABCA13 top pos corr
# ABCG2 topneg corr


# Selecting top associated genes for categorical covariates.

# Removing white spaces in the variable values.

covariates_data_filt$sex <- trimws(covariates_data_filt$sex, which = "left")

# Replacing unknown by NA.

covariates_data_filt$sex[covariates_data_filt$sex == "unknown"] <- NA

table(covariates_data_filt$sex)

t_tests <- c()

for (i  in 1:nrow(gene_expression)){
  
  test_results <- t.test(as.numeric(gene_expression[i,]) ~ covariates_data_filt[,4])
  
  t_tests <- c(t_tests,test_results$statistic)
}


rownames(gene_expression)[which.max(t_tests)]

max(t_tests,na.rm = TRUE)

rownames(gene_expression)[which.min(t_tests)]

min(t_tests,na.rm = TRUE)

# ABHD16B top pos corr
# ABHD5 topneg corr


# Selecting top associated genes for categorical covariates.

# Genes associated with mechanical ventilation.

table(covariates_data_filt$mechanical_ventilation)

t_tests <- c()

for (i  in 1:nrow(gene_expression)){
  
  test_results <- t.test(as.numeric(gene_expression[i,]) ~ covariates_data_filt[,5])
  
  t_tests <- c(t_tests,test_results$statistic)
}


rownames(gene_expression)[which.max(t_tests)]

max(t_tests,na.rm = TRUE)

rownames(gene_expression)[which.min(t_tests)]

min(t_tests,na.rm = TRUE)

# ABCD4 top pos t-test statistic
# ABHD5 top neg t-test statistic


# Selecting top associated genes for categorical covariates.

# Genes associated with disease status

table(covariates_data_filt$disease_status)

t_tests <- c()

for (i  in 1:nrow(gene_expression)){
  
  test_results <- t.test(as.numeric(gene_expression[i,]) ~ covariates_data_filt[,6])
  
  t_tests <- c(t_tests,test_results$statistic)
}


rownames(gene_expression)[which.max(t_tests)]

max(t_tests,na.rm = TRUE)

rownames(gene_expression)[which.min(t_tests)]

min(t_tests,na.rm = TRUE)

# ABCB6 top pos t-test statistic
# ABHD17A top neg t-test statistic
```

In this section I will format the column names of the filtered covariates dataframe.

```{r}
# Renaming columns of the covariate dataframe and claining data.

colnames(covariates_data_filt) <- c("HFD-45","Ferritin (ng/mL)",

          "Fibrinogen (mg/dL)","Sex","Mechanical Ventilation", "Disease Status")

# Replacing unknown values by NAs.

covariates_data_filt$`Fibrinogen (mg/dL)`[covariates_data_filt$`Fibrinogen (mg/dL)` == "unknown"] <- NA

covariates_data_filt$`Fibrinogen (mg/dL)` <- as.numeric(covariates_data_filt$`Fibrinogen (mg/dL)`)

covariates_data_filt$`Ferritin (ng/mL)`[covariates_data_filt$`Ferritin (ng/mL)` == "unknown"] <- NA

covariates_data_filt$`Ferritin (ng/mL)` <- as.numeric(covariates_data_filt$`Ferritin (ng/mL)`)
```
I created a summary statistics table for our continuous and categorical data stratifying by Sex. To do that I first formated the covariates data. Then using the tableone palcage I created the tables that included means and standard deviations for continuous variables and percentages for categorical ones.

```{r, warning=FALSE}
library(tableone)

# Formatting the values.

covariates_data_filt$Sex[covariates_data_filt$Sex == "male"] <- "Male"

covariates_data_filt$Sex[covariates_data_filt$Sex == "female"] <- "Female"

covariates_data_filt$`Disease Status`[covariates_data_filt$`Disease Status` == "disease state: COVID-19"] <- "COVID-19"

covariates_data_filt$`Disease Status`[covariates_data_filt$`Disease Status` == "disease state: non-COVID-19"] <- "non-COVID-19"


# Using the package tableone to create the table.

tab <- CreateTableOne(vars = colnames(covariates_data_filt),
                      strata = "Sex",
                      data = covariates_data_filt,
                      factorVars = c("Sex","Mechanical Ventilation", "Disease Status"))


print(tab, showAllLevels = TRUE, quote = FALSE, noSpaces = TRUE, test = FALSE)

# Using the package xtable to create the summary statistics table in latex format.

library(xtable)

tab_latex <- print(tab, showAllLevels = TRUE, quote = FALSE, noSpaces = TRUE, test = FALSE, printToggle = FALSE, format = "latex")    

print(xtable(tab_latex),type = "latex",)
```

In this section I define the plotting functions 

```{r}
# Merging gene expression and covariate data.frames

selected_gene_tran <- data.frame(t(as.matrix(gene_expression)))

data <- data.frame(covariates_data_filt,selected_gene_tran,check.names = FALSE)
```

# Generating plots for all selected covariates.

This section contains the code to generate the histograms, scatteplots and boxplots for all selected covariates and each selected gene.

```{r,eval = FALSE}
#  Dividing continous and categorical covariates into two data.frames

cont_covariate_list <- colnames(covariates_data_filt)[c(1,2,3)]

cat_covariate_list <-  colnames(covariates_data_filt)[c(4,5,6)]

# Writing function to create all plots.


function_vui <- function(d_f,list_of_genes,list_cont_cov,list_cat_cov){
  
  for (gene in list_of_genes){
    
    # Plotting histograms
    
    plot <- ggplot(d_f, aes(x = !!sym(gene))) + geom_histogram() + 
    labs(title = paste("Histogram for gene: ", gene, sep ="")) + theme_minimal() +
    labs( y = "Counts")
    print(plot)
    
    # Plotting scatterplot.
    
    for (cont_cov in list_cont_cov){
      
      plot_2 <- ggplot(d_f, aes(y = !!sym(gene),
    x = !!sym(cont_cov))) +
    geom_point() +
    ggtitle(paste("Expression of ",
    gene," by Hospital Free Days", sep = "")) + theme_minimal()
    
    print(plot_2)
      
    }
    
    # Plotting boxplots.
    
    for (cat_cov in list_cat_cov){
      
        # Removing potential NA categories in the variable
      
        data_filt <- d_f[!is.na(d_f[,cat_cov]),]
      
        plot_3 <- ggplot(data_filt, aes(x = !!sym(cat_cov) ,
                                        y = !!sym(gene),fill = !!sym(cat_cov))) +
        geom_boxplot(na.rm = TRUE) + ggtitle(paste("Expression of ",
        gene," by ", cat_cov, sep = "")) + theme_minimal()
        
        print(plot_3)
    
      }

    }

}

library(ggplot2)

selected_genes <- c("ABCD4", "ABCB6", "A4GALT",
                    "AAMP", "ABCA13", "ABCG2", "ABHD16B", "ABHD5", "ABHD17A","A1BG")

# Generating plots for all the genes

function_vui(data,selected_genes,cont_covariate_list,cat_covariate_list)
```

# Plots for the final assignment.

This section contains the function to generate the plots for the final submission of the final assignment.
```{r}

# Creating function to plot histogram.

function_histogram_vui <- function(d_f,gene_name){
        
    plot <- ggplot(d_f, aes(x = !!sym(gene_name))) + 
      
      geom_histogram(bins = 30, fill = "steelblue", 
    
    color = "black") + 
      
    labs(title = paste("Histogram for gene: ", gene_name, sep =""), 
         
         y = "Counts", x = paste0(gene_name, " Expression") ) +
      
    theme_classic() 
      
    print(plot)
    
    return(plot)
}

# Running the function

plot_hist <- function_histogram_vui(data,"ABCD4")

# Saving plot in png format using ggsave
ggsave("C:/Users/Guest1/Dropbox/VuiFinal/Plot_ABCD4_Hist.png", plot = plot_hist, width = 6, height = 4, dpi = 300)
```
Here the function that creates the final scatter plot is created and executed.

```{r}
function_scatter_vui <- function(d_f, gene_name, cont_cov) {
  
  plot <- ggplot(d_f, aes(y = !!sym(gene_name),
                          
                          x = !!sym(cont_cov))) +
    geom_point(color = "steelblue", size = 3) +
    
    labs(title = paste(gene_name, "Expression Vs ",
                       
                       cont_cov),y = paste0(gene_name, " Expression")) +
    
    theme_classic()
  
  print(plot)
  
  return(plot)
}


# Running the function

dataplot_scat <- function_scatter_vui(data,"ABCD4","HFD-45")

ggsave("C:/Users/Guest1/Dropbox/VuiFinal/Plot_ABCD4_Scatter.png", 
       plot = dataplot_scat, width = 6, height = 4, dpi = 300)
```

Here the function that stratifies the boxplot by two different covariates is defined.

```{r}

function_box <- function(d_f,gene_name,cat_cov_1,cat_cov_2){
  
        # Filtering out data with missing values.
    
        data_filt <- d_f[!(is.na(d_f[,cat_cov_1]) | is.na(d_f[,cat_cov_2])),]
      
        plot <- ggplot(data_filt, aes(x = !!sym(cat_cov_1), y = !!sym(gene_name), fill = !!sym(cat_cov_2))) +
          
       geom_boxplot(position = position_dodge(width = 0.8)) +
          
       labs(x = cat_cov_1, y = paste0(gene_name, " Expression"), fill = cat_cov_2) +
          
       theme_classic()
        
       print(plot)
       
       return(plot)
}

dataplot_box <- function_box(data,"ABCD4","Sex","Disease Status")

ggsave("C:/Users/Guest1/Dropbox/VuiFinal/Plot_ABCD4_Box.png", 
       
       plot = dataplot_box, width = 6, height = 4, dpi = 300)
```

In this section we will create a heatmap with annotations. We selected the 9 genes that presented the top positive and negative associations with our selected covariates.

```{r}
genes_selected_for_heatmap <- c("ABCD4", "ABCB6", "A4GALT", "AAMP", "ABCA13", 
                                "ABCG2", "ABHD16B", "ABHD5", "ABHD17A","A1BG")


expression_selected <- data[,genes_selected_for_heatmap]

rownames(expression_selected) <- paste("Sample",seq(1,nrow(expression_selected)))

covariates_data <- data[,c(4,6)]

rownames(covariates_data) <- paste("Sample",seq(1,nrow(expression_selected)))

library(pheatmap)

# Sving Heatmap.

pheatmap(t(expression_selected),annotation_col = covariates_data, scale = "row", 
         color = colorRampPalette(c("green", "white", "red"))(50),fontsize = 5, filename = "C:/Users/Guest1/Dropbox/VuiFinal/Heatmap.png",
         show_colnames = FALSE, main = "Heatmap of 10 selected genes")
```

```{r}
# Plotting heatmap.

pheatmap(t(expression_selected),annotation_col = covariates_data, scale = "row", 
         color = colorRampPalette(c("green", "white", "red"))(50),fontsize = 5,
         show_colnames = FALSE, main = "Heatmap of 10 selected genes")
```

# Principal component analysis.

For the new plot we are going to generate a principal component analysis of the data.

```{r}
# Getting gene expression.

gene_expression <- data[,seq(7,ncol(data))]

gene_expression <- gene_expression[, colSums(gene_expression != 0) > 0]

# Getting covariates.

covariates_data <- data[,seq(1,6)]

# Computing principal components of gene expression data using prcomp.

pca_result <- prcomp(gene_expression, center = TRUE, scale. = TRUE)

pca_scores <- as.data.frame(pca_result$x)

pca_scores$Sex <- covariates_data$Sex

pca_scores$DiseaseStatus <- covariates_data$`Disease Status`


library(ggplot2)

# Plotting PCA results using ggplot2.

pca_plot <- ggplot(pca_scores, aes(x = PC1, y = PC2, 
                                   
                                   color = DiseaseStatus, shape = Sex )) +
  geom_point(size = 3) +
  
  theme_classic() +
  
  labs(title = "PCA",
       
   x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
  
   y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)"))


ggsave("C:/Users/Guest1/Dropbox/VuiFinal/PlotPCA.png", plot = pca_plot, width = 6, height = 4, dpi = 300)

pca_plot
```

